\mythesistitleNOTSYNPOSIS

### 2012-01-06 src/0021ded.tex
Chapter 2 edited. See diff for details. Too many to remember. or maybe too little.
Added more stuff to chapter 2. see diff.

{
  \fontencoding{T1}
%  \fontfamily{cmdh} 
  \fontfamily{bch} %{put}  %{bch} % {pbk}
%  \fontfamily{pag} %
  \selectfont%
  {\large \noindent The process of preparing programs for the digital
    computer is especially attractive, not only because it can be
    economically and scientifically rewarding, but also because it can
    be an aesthetic experience much like composing poetry or
    music.%
    {\small \par \hfill \emph{The Art of Computer Programming},
      D.\;E.\;K.}%
  }%
  \vskip 5ex%
  {\Large \noindent The rhetoric and mythos of science create the
    comforting image of linear progression toward truth.%
    {\normalsize \par \hfill \emph{New Oxford American
        Dictionary, 2nd Ed.}}%\textbf{mythos};
  }%  
  \vskip 5ex%
  {\huge%
    \noindent Art is the solution to chaos. \par %
  }%
}

>>>>

{
  \fontencoding{T1}
%  \fontfamily{cmdh} 
  \fontfamily{bch} %{put}  %{bch} % {pbk}
%  \fontfamily{pag} %
  \selectfont%
  {\large \noindent The process of preparing programs for the digital
    computer is especially attractive, not only because it can be
    economically and scientifically rewarding, but also because it can
    be an aesthetic experience much like composing poetry or
    music.%
    {\small \par \hfill \emph{The Art of Computer Programming},
      D.\;E.\;K.}%
  }%
  \vskip 5ex%
  {\large \noindent The rhetoric and mythos of science create the
    comforting image of linear progression toward truth.%
    {\normalsize \par \hfill \emph{New Oxford American
        Dictionary, 2nd Ed.}}%\textbf{mythos};
  }%  
  \vskip 15ex%
  {\large%
    \noindent Art is the solution to chaos. \par %
    {\normalsize \par \hfill \emph{citation needed.}}
  }%
}

### 2012-01-06 src/010chap1.tex
Added more stuff to chapter 2. see diff.

\section{Basic preliminaries - general definitions and nomenclature}
\label{sec:basicprelim}
\tnote{(definitions theorems etc) needed if any - graph theory}

\begin{definition}{\emph{\Cop}}
  \tnote[important]{define cop here. refer to it in chap 2 or
    duplicate it using slash def }
  \begin{enumerate}
  \item \Cop
  \item Strong \cop
  \item \Cop order or \cop permutation
  \end{enumerate}
\end{definition}

\begin{figure}[t] %[htbp]
  \centering
  {\figtabsize
    \begin{tabular}[h]{l|lcccl}
      $M_1$: & $M_1'$: &&&& $M_2$:\\
      &&&&&\\
      \begin{tabular}[h]{llll}
        $c_1$ & $c_2$ &$c_3$ &$c_4$\\
        &&&\\
        \un & 0   & \un & 0\\
        0   & \un & 0   & \un \\
        \un & 0   & 0   & \un
      \end{tabular}
      &
      \begin{tabular}[h]{llll}
        $c_3$ &$c_1$ &$c_4$& $c_2$\\
        &&&\\
        \un & \un & 0 & 0\\
        0 & 0 & \un & \un \\
        0 & \un & \un & 0
      \end{tabular}
      &&&&
      \begin{tabular}[h]{llll}
        $d_1$ & $d_2$ &$d_3$ &$d_4$\\
        % &&&\\
        &&&\\
        \un & \un & 0 & 0\\
        0 & \un & \un & 0 \\
        0 & \un & 0 & \un 
      \end{tabular}
    \end{tabular}
  }

  \caption[\figtabsize Matrices with and without COP.]{\figtabsize
    Matrices with and without COP. $M_1$ has COP because by permuting
    its columns, $c_1$-$c_4$, one can obtain $M_1'$ where the {\un}s
    in each row are consecutive. $M_2$, however, does not have COP
    since no permutation of its columns, $d_1$-$d_4$, will arrange
    {\un}s in each row consecutively \cite{d08phd}.}

  \label{fig:cop-matrix}
\end{figure}

\begin{definition}{\emph{\Crop}}
  \tnote[important]{define circ1 here. refer to it in chap 2 or
    duplicate it using slash def }
\end{definition}

\begin{definition}{\emph{Binary matrix}}
  Let $M$ be an $n \times m$ matrix. $m_{i,j}$ denotes its $(i,j)$th
  element, i.e. element at $i$th row and $j$th column. $M$ is a binary
  matrix if each of its element is 0 or \un.
  In other words, for all $i \in [n]$ and $j \in [m]$, $m_{i,j} \in
  \{0, \un\}$ 
\end{definition}

>>>>

\section{Basic preliminaries - general definitions and nomenclature}
\label{sec:basicprelim}

\begin{definition}[\emph{Binary matrix}]
  Let $M$ be an $n \times m$ matrix. $m_{i,j}$ denotes its $(i,j)$th
  element, i.e. element at $i$th row and $j$th column. $M$ is a binary
  matrix if each of its element is 0 or \un.
  In other words, for all $i \in [n]$ and $j \in [m]$, $m_{i,j} \in
  \{0, \un\}$ 
\end{definition}

\begin{definition}[\emph{set system}]
\end{definition}

\begin{definition}[\emph{\Cop}]
  \tnote[important]{define cop here. refer to it in chap 2 or
    duplicate it using slash def }
  \begin{enumerate}
  \item \Cop
  \item Strong \cop
  \item \Cop order or \cop permutation
  \item \Cop for set systems\tnote{see mcc04}
  \end{enumerate}
\end{definition}

\figcopmatrix
\tnote[important]{have a figure for set systems also}

\begin{definition}[\emph{\Crop}]
  \tnote[important]{define circ1 here. refer to it in chap 2 or
    duplicate it using slash def }
\end{definition}

\begin{definition}[\emph{Graph, Tree}]

\end{definition}

\begin{definition}[\emph{Maximal clique}]

\end{definition}

### 2012-01-06 src/020chap2.tex
Added more stuff to chapter 2. see diff.

\tnote{DEFINE COP SOMEWHERE}

@@

% \begin{definition}{\emph{Matrices that define
%       graphs.\cite[Def.~2.4]{d08phd}}} %
\begin{definition}
  {\em Matrices that define graphs. \cite[Def.~2.4]{d08phd}.}  
  \label{def:graphmatrices} 
  Let $G$ and $H$ be defined as follows. $G = (V,E_G)$ is a graph with
  vertex set $V = \{v_i \mid i \in [n]\}$ and edge set $E_G \subseteq
  \{(v_i,v_j) \mid i, j \in [n]\}$ such that $|E_G| = m$. $H = (A, B,
  E_H)$ is a bipartite graph with partitions $A = \{a_i \mid i \in
  [n_a]\}$ and $B = \{b_i \mid i \in [n_b]\}$.
  \begin{enumerate}[{\ref{def:graphmatrices}}--i.]
  \item \emph{Adjacency matrix} of $G$ is the symmetric $n \times n$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(v_i,v_j) \in E_G$
    for all $i,j \in [n]$.
  \item \emph{Augmented adjacency matrix} of $G$ is obtained from its
    adjacency matrix by setting all main diagonal elements to \un, \ie
    $m_{i,i} = \un$ for all $i \in [n]$.
  \item \emph{Maximal clique matrix} or \emph{vertex-clique incidence
      matrix} of $G$ is the $n \times k$ binary matrix $M$ with
    $m_{i,j} = \un$ \iff $v_i \in C_j$ for all $i \in [n], j \in [k]$
    where $\{C_j \mid j \in [k]\}$ is the set of maximal cliques of
    $G$.
  \item \emph{Half adjacency matrix} of $H$ is the $n_a \times n_b$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(a_i, b_j) \in E_H$.
  \end{enumerate}
\end{definition}

@@

  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote{cite BHY00}
    \tnote[minor]{add CROP to glossary}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote{cite Ben59, Haj57}.
    $G$ is an interval graph \iff its maximal clique matrix has COP
    \cite{fg65}\endnote{This follows \cite{gh64} which states that the
      maximal cliques of interval graph $G$ can be linearly ordered
      \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.}
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.\endnotemark[2]
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.\endnote{The
        set of unit interval graphs and the set of proper interval
        graphs are the same} \tnote{cite {rob69,gar07} in endnote. pg
        33 dom}
    \end{enumerate}
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item \label{def::convexbi} $H$ is \emph{convex bipartite on columns (rows)} if its half adjacency matrix
    has \COP on rows (columns).
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

@@

\clearpage      % Print pending floats

@@

\subsection{Tucker's forbidden submatrices for \COP}

@@

\temptext{
  \begin{enumerate}[-I-]
  \item PQ trees. Define. Concise flowchart of algorithm? [BL76]\\
    The construction of PQ trees in \cite{bl76} draws on the close
    relationship of matrices with COP to interval graphs. A PQ tree of
    a matrix is one that stores all row (column) permutations of the
    matrix that give the COP orders (there could be multiple orders of
    rows or columns) of the matrix. This is constructed using an
    elaborate linear time procedure and is also a test for
    planarity\tnote{check check check. both interval graph and
      planarity in this paper?}.
  \item Variations - PQR, PC etc. [MM96, Hsu01, McC04]\\
    PQR trees is a generalized data structure based on PQ trees
    \cite{mm96,mpt98}.  \cite{tm05} describes an improved algorithm to
    build PQR
    trees. \tnote{improv in terms of what?}.\\
    Generalized PQ trees McC04 - brief. Will be elaborated in
    Section~\ref{sec:surveycertalgo}\\
    Hsu also invented PC trees \cite{wlh01}\tnote{This result first
      appeared inproc ISAAC92} which is claimed to be much easier to
    implement.
  \item Set theoretic characterizations [Hsu02, NS09]\\
    \cite{wlh02} describes the simpler algorithm for
    COT.\tnote[pressing]{in terms of what?}\\
    \cite{nsnrs09} describes a characterization of consecutive-ones
    property solely based on the cardinality properties of the set
    representations of the columns (rows); every column (row) is
    equivalent to a set that has the row (column) indices of the rows
    (columns) that have one entries in this column (row). This is
    interesting and relevant, especially to this thesis because it
    simplifies COT to a great degree by reducing the solution search
    space owing to the a simple set theoretic characterization.\\
  \item Parallel algorithms [AS95, BS03, CY91]\\
  \item
  \end{enumerate}

@@

\clearpage      % Print pending floats
\section{Optimization problems in COP}
\label{sec:surveycopopt}
\tnote{Expand on ref:sec:optcop}

>>>>

@@

% \begin{definition}{\emph{Matrices that define
%       graphs.\cite[Def.~2.4]{d08phd}}} %
\begin{definition}
  {\em Matrices that define graphs. \cite[Def.~2.4]{d08phd}.}  
  \label{def:graphmatrices} 
  Let $G$ and $H$ be defined as follows. $G = (V,E_G)$ is a graph with
  vertex set $V = \{v_i \mid i \in [n]\}$ and edge set $E_G \subseteq
  \{(v_i,v_j) \mid i, j \in [n]\}$ such that $|E_G| = m$. $H = (A, B,
  E_H)$ is a bipartite graph with partitions $A = \{a_i \mid i \in
  [n_a]\}$ and $B = \{b_i \mid i \in [n_b]\}$.
  \begin{enumerate}[{\ref{def:graphmatrices}}--i.]
  \item \emph{Adjacency matrix} of $G$ is the symmetric $n \times n$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(v_i,v_j) \in E_G$
    for all $i,j \in [n]$.
  \item \emph{Augmented adjacency matrix} of $G$ is obtained from its
    adjacency matrix by setting all main diagonal elements to \un, \ie
    $m_{i,i} = \un$ for all $i \in [n]$.
  \item \emph{Maximal clique matrix} or \emph{vertex-clique incidence
      matrix} of $G$ is the $n \times k$ binary matrix $M$ with
    $m_{i,j} = \un$ \iff $v_i \in C_j$ for all $i \in [n], j \in [k]$
    where $\{C_j \mid j \in [k]\}$ is the set of maximal cliques of
    $G$.
    \label{def::maxcliquematrix}
  \item \emph{Half adjacency matrix} of $H$ is the $n_a \times n_b$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(a_i, b_j) \in E_H$.
  \end{enumerate}
\end{definition}

@@

  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote{cite BHY00}
    \tnote[minor]{add CROP to glossary}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote{cite Ben59, Haj57}.
    $G$ is an interval graph \iff its maximal clique matrix has COP
    \cite{fg65}\endnote{This follows \cite{gh64} which states that the
      maximal cliques of interval graph $G$ can be linearly ordered
      \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.}

    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.\endnotemark[2]
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.\endnote{The
        set of unit interval graphs and the set of proper interval
        graphs are the same} \tnote{cite {rob69,gar07} in endnote. pg
        33 dom}
    \end{enumerate}
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item $H$ is \emph{convex bipartite on columns (rows)} if its half
    adjacency matrix has \COP on rows (columns).%
    \label{def::convexbi}
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

@@

%\clearpage      % Print pending floats

@@

%\clearpage
\subsection{Tucker's forbidden submatrices for \COP}

@@

%\clearpage
\subsection{$PQ$ tree -- a linear COT algorithm}

Booth and Lueker in their paper \cite{bl76} gave the first linear
algorithm for \cop testing by way of an interval graph recognition
algorithm. Their result has close relations to the characterization of
interval graphs by \cite{gh64}. A graph $G$ is an interval graph \iff
all its maximal cliques can be linearly ordered \stt for any vertex
$v$ in $G$, all the cliques that $v$ is incident on are consecutive in
this order. Clearly, this means that the maximal clique incidence
matrix\footnote{
  Definition~\ref{def:graphmatrices}~\ref{def::maxcliquematrix}} must
have \COP on rows. This algorithm constructs a data structure called
\PQtree. A \PQtree represents all the \COP orderings of the matrix it
is associated with. \cite{bl76}'s algorithm uses the fact that if a
matrix has \COP, a \PQtree for it can be constructed. It is
interesting to note that aside from interval graph recognition and
\COP testing, \PQtree is also useful in other applications like
finding planar embeddings of planar graphs \cite{lec67,mcc04} and
recognizing \CROP in a matrix.

\begin{definition}[\emph{\PQtree \cite{bl76, mcc04}}]\\
  A \PQtree of matrix $M$ with \COP, is a tree with the following properties.
  \begin{enumerate}[i.]
    \singlespacing
  \item Each leaf uniquely represents a row (column) of $M$. The leaf
    order of the tree gives a \COP order for column (row)\endnote{Note
      that \COP order for column requires permutation of rows and vice
      versa.} for $M$.
  \item Every non-leaf node in the tree are labeled $P$ or $Q$.
  \item \label{def::nodep} The children of $P$ nodes are
    unordered. They can be permuted in any fashion to obtain a new
    \COP order for $M$.
  \item \label{def::nodeq} The children of $Q$ nodes are linearly
    ordered. Their order can be reversed to get obtain a new \COP
    order for $M$.
  \end{enumerate}
  \label{def:pqtree}
\end{definition}

Thus, effectively, there exists a bijection between set of matrices
with \COP and the set of \PQtrees (accurately speaking, each matrix
with \COP bijectively maps to an equivalence class of \PQtrees due to
properties (\ref{def::nodep}) and (\ref{def::nodeq})). See
Figure~\ref{fig:pqtree} for an example of \PQtree.

\begin{figure}[htbp]
  \centering
  \begin{tabular}[h]{lr}
    \begin{tabular}[h]
      {>{\columncolor{\tblhcolor}} l lllllllllll|}
      \hline
      $r_1$   &\un &0   &\un &0   &0   &0    &0   \\
      $r_2$   &\un &0   &\un &0   &0   &0    &0   \\
      $r_3$   &\un &0   &0   &\un &0   &0    &0   \\
      $r_4$   &\un &0   &0   &\un &0   &0    &0   \\
      $r_5$   &\un &0   &0   &0   &0   &0    &0   \\
      $r_6$   &\un &\un &0   &0   &0   &0    &0   \\
      $r_7$   &0   &\un &0   &0   &0   &\un  &0   \\
      $r_8$   &0   &\un &0   &0   &\un &\un  &0   \\
      $r_9$   &0   &\un &0   &0   &\un &\un  &0   \\
      $r_{10}$ &0   &\un &0   &0   &\un &\un  &\un \\
      $r_{11}$ &0   &\un &0   &0   &\un &0    &\un \\
      \hline
    \end{tabular}
    &
    \includegraphics[scale=0.3]{../img/pqtree.png}

  \end{tabular}

  \caption[\figtabsize An example for \PQtree]{\figtabsize An example for \PQtree.
    Permuting the order of the left child of the root, we see that
    \temptext{(d,a,b,c,e,f,g,h,i,j,k)} is a \COP order. Reversing the
    order of the right child of the root, we see that
    \temptext{(a,b,c,d,e,f,k,j,h,i,g) } is yet another \COP order. \temptext{PLACEHOLDER IMG.}
    \cite[Fig.~1\endnote{\cite{mcc04}illustrates \COP on rows. Our
      convention in this document is \COP on columns and thus example
      matrix has been transposed. }]{mcc04}}
  \label{fig:pqtree}
\end{figure}

The \cite{bl76} algorithm with input $n \times m$ matrix $M$ starts
with a \PQtree for a vacuous $n \times 0$ matrix $M'$ (submatrix
induced by 0 columns). This is known as a {\em universal} \PQtree
which is one with its root as a $P$ node and only leaves as its
children -- each leaf representative of a row of input (by definition
of \COP for columns). This induced submatrix $M'$ vacuously has \COP.
Each column is then added iteratively to $M'$ to check if the new $M'$
has \COP.  By a complicated, but linear, procedure the algorithm does
one of the following actions in each iteration: (a) declare that $M$
has no \COP, or (b) modify the current \PQtree to represent the new
$M'$ (which clearly, must have \COP, since if not, option (a) would
have been executed).

\temptext{
  \begin{enumerate}[ -I-]
  \item  Variations - PQR, PC etc. [MM96, Hsu01, McC04]\\
    PQR trees is a generalized data structure based on PQ trees
    \cite{mm96,mpt98}.  \cite{tm05} describes an improved algorithm to
    build PQR
    trees. \tnote{improv in terms of what?}.\\
    Generalized PQ trees McC04 - brief. Will be elaborated in
    Section~\ref{sec:surveycertalgo}\\
    Hsu also invented PC trees \cite{wlh01}\tnote{This result first
      appeared inproc ISAAC92} which is claimed to be much easier to
    implement.
  \item Set theoretic characterizations [Hsu02, NS09]\\
    \cite{wlh02} describes the simpler algorithm for
    COT.\tnote[pressing]{in terms of what?}\\
    \cite{nsnrs09} describes a characterization of consecutive-ones
    property solely based on the cardinality properties of the set
    representations of the columns (rows); every column (row) is
    equivalent to a set that has the row (column) indices of the rows
    (columns) that have one entries in this column (row). This is
    interesting and relevant, especially to this thesis because it
    simplifies COT to a great degree by reducing the solution search
    space owing to the a simple set theoretic characterization.\\
  \item Parallel algorithms [AS95, BS03, CY91]
  \end{enumerate}

@@

%\clearpage      % Print pending floats
\section{Optimization problems in COP}
\label{sec:surveycopopt}
\tnote{Expand on ref:sec:optcop}
